---
title: "Zoo955 - Raster Data"
author: "Hilary Dugan"
date: "February 20, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## What is a Raster? (from raster package)
A raster is a spatial (geographic) data structure that divides a region into rectangles called 'cells' (or
'pixels') 

* Each cell has 1 or more data values

Such a data structure is also referred to as a 'grid' and is often contrasted with 'vector' data that is used to represent points, lines, and polygons.

#### Raster advantages & disadvantages over vector data

* Raster data usually requires more storage space than vector data 
* Raster data can be more difficult to re-project and scale
* Raster data can be more difficult to update 

* Raster data can be computationally less expensive to render graphically
* Easy to plot 

## Types of Raster Data

* Climate data
* Land cover data
* Elevation data

Many of these are satellite derived, and the resolution depends on the satellite resolution

## How is raster data stored? 
There are many ways raster data is stored. Each file type is specialized for its use. 

-------------------------------------------------------------
Type              Used for/by                       Format 
----------------- ---------------------- ----------------- 
ARC Digitized     National Geospatial    .img or .ovr
Raster Graphics   Intelligence Agency
(ADRG)                 

JPEG2000          Compression technique  .jp2, .j2c, .j2k, or .jpx
                  for large imagery

GeoTiff           Georeferencing within  .tif, .tiff, or .tff
                  a TIFF file

NetCDF            climatology            .nc
Network Common    meteorology
Data Form         oceanography
-------------------------------------------------------------

Table: Some common raster file types
  
You don't need to know the file type details, but you will become familiar with with formats can be read into R

## Using raster data in R
The R Raster package: https://cran.r-project.org/web/packages/raster/vignettes/Raster.pdf

"A notable feature of the raster package is that it can work with raster
datasets that are stored on disk and are too large to be loaded into memory
(RAM). The package can work with large files because the objects it creates
from these files only contain information about the structure of the data, such
as the number of rows and columns, the spatial extent, and the filename, but it
does not attempt to read all the cell values in memory. In computations with
these objects, data is processed in chunks."


## Data structures 
#### Raster Layer
A RasterLayer object represents single-layer (variable) raster data

#### Raster Stack vs. Brick

* There are two classes for multi-layer data the __RasterStack__ and the __RasterBrick__ 
* __RasterStack__ can be formed from separate files and/or from a few layers ('bands') 
    * RasterStack is a collection of RasterLayer objects with the same
spatial extent and resolution
    * Basically a list of RasterLayer objects
* __RasterBrick__ can only be linked to a single (multi-layer) file
    * A multilayered object, but can only refer to a single file 
    * More effecient processing 

## Import data
The raster package makes life easy for importing raster data. I have provided two different storage formats for Lake Mendota NLCD data. 

1) .grd = native grid format for raster package
2) .tif = geoTIFF

```{r nlcd, message=FALSE, warning=FALSE}
library(raster,verbose = F)
nlcd = raster('Data/MendotaNLCD.tif')
nlcd = raster('Data/MendotaNLCD.grd')
nlcd
```

Can explore the data values
```{r, message=FALSE, warning=FALSE}
hist(nlcd)
```

Can plot the data
```{r, message=FALSE, warning=FALSE}
plot(nlcd)
image(nlcd)
```

Look at the file structure
```{r, message=FALSE, warning=FALSE}
str(nlcd)
```

Notice that there are attributes and a legend
```{r, message=FALSE, warning=FALSE}
atts = nlcd@data@attributes[[1]]
head(atts,20)
nlcd@legend
```