---
title: "Zoo 955 - Introduction to Zoo 955"
author: "Hilary Dugan"
date: "January 23, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/',echo = TRUE)
```

# 955 basics

* Class notes: https://github.com/hdugan/Zoo955
* Class ZOOM: https://zoom.us/j/726741289

# 955 basics 

* R verison >= 3.4 
* Recommend RStudio as a code editor 
    * This is what I will use in class 
    * RStudio Desktop 1.1.414 
* Your primary reference will probably be community answers on Stack Overflow (https://stackoverflow.com)
* Recommend creating a username and upvoting answers you find helpful

# 955 Assesment 

* Grades will be based on:
    * short assignments (25%)
    * participating in class discussion (50%) 
    * final project based on individual research topics that will be presented during the final two weeks of the semester (25%)
* Numerical grades are assigned as follows: 93-100 (A), 88-92 (AB), 82-87 (B), 78-81 (BC).

* There will be short weekly assignments 
* Typically I'll ask that you email me your assignments as a pdf 
    * R markdown tutorial 
    * https://ismayc.github.io/rbasics-book/4-rmarkdown.html
    * Or however you'd like to make a pdf

* There's no class Feb 13th. Instead, short proposals of your intended research endeavour will be due. 

# The spatial basics 

* Commonly used R packages 
    * `library(rgdal)`
    * `library(rgeos)`
    * `library(sp)`
    * `library(maps)`
    * `library(raster)`
    * `library(maptools)`

# Why use R for spatial analysis?
Pros:

* Breadth of statistical packages
* Integration with other data/stats
* Free, easy to access

Cons: 

* Memory limitations 

# Spatial Objects
  |       |  Without attributes | With attributes (data) |
  | ----- | ------------------  | --------------  |
  |Points |  SpatialPoints      | SpatialPointsDataFrame|
  |Lines |  SpatialLines      | SpatialLinesDataFrame|
  |Polygons |  SpatialPolygons      | SpatialPolygonsDataFrame|
  |Raster |  SpatialGrid      | SpatialGridDataFrame|
  |Raster |  SpatialPixels      | SpatialPixelsDataFrame|
  
# Spatial objects
* Spatial class has two slots
* 1) bounding box, dimension of the plot
* 2) CRS class object defining the coordinate
reference system, and may be set to CRS(as.character(NA)),
its default value.
* Operations on `Spatial` objects should update or copy these
values to the new Spatial objects being created  
  

# SpatialPoints
* The most basic spatial data object is a point, which may have
2 or 3 dimensions
* A single coordinate, or a set of such coordinates, may be used
to define a SpatialPoints object
* The points in a SpatialPoints object may be associated with
a row of attributes to create a SpatialPointsDataFrame object

# Example: SpatialPoints
```{r spoint, echo=T, eval=T}
library(sp)
spoint <- data.frame(y = 43.07, x = -89.40)
summary(spoint)
str(spoint)
coords <- SpatialPoints(spoint)
str(coords)
summary(coords)
```

# SpatialPointsDataFrame
Add data to make a SpatialPointsDataFrame object. 

```{r sdata, echo=T, eval=T}
# library(sp)
sdata <- SpatialPointsDataFrame(coords, data = data.frame(population = 327000))
str(sdata)
summary(sdata)
```

# Spatial Points + Data Diagram 
![*Figure credit: Roger S. Bivand, Edzer J. Pebesma and Virgilio G?mez-Rubio. 
Applied Spatial Data Analysis with R 
Springer. August 2008.*](Figs/SpatialPointsDiagram.PNG)


# Spatial Lines & Polygons
* A Line object is a collection of 2D coordinates
* A Polygon object is a Line object with equal first and last
coordinates
* A Lines object is a list of Line objects, such as all the
contours at a single elevation; the same relationship holds
between a Polygons object and a list of Polygon objects, such
as islands belonging to the same county
* SpatialLines and SpatialPolygons objects are made using
lists of Lines or Polygons objects respectively
* SpatialLinesDataFrame and SpatialPolygonsDataFrame
objects are defined using SpatialLines and SpatialPolygons
objects and standard data frames, and the ID fields are here
required to match the data frame row names

# Example: Spatial Polygons
```{r spoly, echo=T, eval=T}
sdata <- data.frame(y = c(43.07,43.10,43.08,43.07), 
                    x = c(-89.40,-89.41,-89.5,089.40))
str(sdata)
spoly <- Polygons(list(Polygon(sdata)), ID = "madison")
summary(spoly)
spolys <- SpatialPolygons(list(spoly))
summary(spolys)
```

# Spatial Polygons Diagram 
![*Figure credit: Roger S. Bivand, Edzer J. Pebesma and Virgilio G?mez-Rubio. 
Applied Spatial Data Analysis with R 
Springer. August 2008.*](Figs/SpatialPolygonsDiagram.PNG)

# Spatial grids and pixels
* There are two representations for data on regular rectangular
grids (oriented N-S, E-W): `SpatialPixels` and `SpatialGrid`
* SpatialPixels are like SpatialPoints objects, but the
coordinates have to be regularly spaced; the coordinates are
stored, as are grid indices
* SpatialPixelsDataFrame objects only store attribute data
where it is present, but need to store the coordinates and grid
indices of those grid cells
* SpatialGridDataFrame objects do not need to store
coordinates, because they fill the entire defined grid, but they
need to store NA values where attribute values are missing

# Spatial Grids Diagram 
![*Figure credit: Roger S. Bivand, Edzer J. Pebesma and Virgilio G?mez-Rubio. 
Applied Spatial Data Analysis with R 
Springer. August 2008.*](Figs/SpatialGridDiagram.PNG)

# Raster Package
Often, instead of using `SpatialGrid`, your prefered obejct format will be a `RasterLayer`

Get to know the Raster package

* Description: Reading, writing, manipulating, analyzing and modeling of gridded spatial data. The packageimplements basic and high-level functions. Processing of very large files is supported

```{r raster, echo=T, eval=T}
library(raster)
f <- system.file("external/test.grd", package="raster")
r <- raster(f)
r
plot(r)
```

Can also have `RasterBrick` objects. Basically multi-layer `RasterLayer` objects. We'll see more of these later in the semester. 
```{r rasterbrick, echo=T, eval=T}
b <- brick(system.file("external/rlogo.grd", package="raster"))
b
```

# Homework 
Make an r markdown pdf. It can be anything you want. Send it to me. 
