---
title: "ZOO955 - Overlays"
author: "Hilary Dugan"
date: "2/22/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vector and Raster Data ##
It is common that you will have to use both vector and raster data in your work. Today we'll explore some common techniques for working with multiple data types. 

#### Lake point data ####
Read in lake point data from .csv 
```{r,message=F,warning=F,fig.width=3,fig.height=3}
library(sp)
library(raster)
library(rgdal)
library(rgeos)
lakes = read.csv('Data/nlakes.csv',stringsAsFactors = F)
coordinates(lakes) = ~Longitude+Latitude
crs(lakes) <- CRS("+proj=longlat +datum=WGS84")
plot(lakes,cex=2)
text(lakes,labels = lakes$Lake,cex=0.8)
```

#### Counties data ####
Read in lake point data from .csv 
```{r,message=F,warning=F}
counties = readOGR('C:/Users/hdugan/Dropbox/RandomR/GIS_Wisconsin/County_Boundaries_24K/County_Boundaries_24K.shp',layer='County_Boundaries_24K')
vilas = counties[counties@data$COUNTY_NAM == 'Vilas',]
plot(vilas)
```

#### National hydrography shapefiles ####
Read in lake point data from .csv 
```{r,message=F,warning=F,fig.keep='last'}
library(rgdal)
vilasNHD = readOGR('Data/vilasNHD.shp',layer = 'vilasNHD')
plot(vilasNHD)
```

#### NLCD raster data ####
Read in lake point data from .csv 
```{r,message=F,warning=F,fig.keep='last'}
nlcd = raster('Data/vilasNLCD.grd')
plot(nlcd)
```

### Cropping rasters ###
You can only crop rasters to be square. So the above raster is already cropped to vilas county. However, if we wanted to be sure:
```{r,message=F,warning=F,error=T,fig.keep='last'}
nlcdvilas = crop(nlcd,vilas)
# Need to make sure they have identical CRS! 
vilas = spTransform(vilas,crs(nlcd)) # transform CRS (usually easier to transform the vector data)
nlcdvilas = crop(nlcd,vilas) # crop the raster the extent of the county
plot(nlcdvilas)
plot(vilas,add = T,lwd=2)
```

### Mask raster ###
Instead we can `mask` the raster. 

`Create a new Raster* object that has the same values as x,`
`except for the cells that are NA (or other maskvalue) in a 'mask'.`

```{r,message=F,warning=F,error=T,fig.keep='last'}
#now use the mask function
nlcdvilas <- mask(nlcd, vilas)
plot(nlcdvilas)
```

### Check if NHD matches up with NLCD ###
Looks like we're dealing with an island issue 
```{r,message=F,warning=F,error=T,fig.keep='last'}
vilas.crs = spTransform(vilasNHD,crs(nlcd))
nlcd.nhd = mask(nlcd,vilas.crs)
plot(nlcd.nhd)
```



### Extract northern study lakes ###
```{r,message=F,warning=F,error=T,fig.keep='last'}
identicalCRS(lakes,vilasNHD)
lakes.crs = spTransform(lakes,crs(vilasNHD))
lakesNHD = crop(vilasNHD,lakes.crs)
plot(lakesNHD)
```

Well that didn't work. This just cropped to the extent. 

Reserve `crop` for raster data sets. Instead we can use `over`. 
```{r,message=F,warning=F,error=T,fig.keep='last'}
b = over(vilasNHD,lakes.crs)
nLakes = vilasNHD[which(!is.na(b)),]
plot(nLakes)
```

### Buffers around study lakes ###
```{r,message=F,warning=F,error=T,fig.keep='last'}
nLakes.m = spTransform(nLakes,crs(nlcd))
buff500 = gBuffer(nLakes.m,width = 100)
plot(buff500)
plot(nLakes.m,add=T,col='lightblue4')
```

### Extract nlcd data 
```{r,message=F,warning=F,error=T,fig.keep='last'}
a = crop(nlcd,buff500)
a = mask(a,buff500)
plot(a)
```

Can ask questions like:

* What is the percent wetlands around these five lakes?
```{r,message=F,warning=F,error=T,fig.keep='last'}
afreq = freq(a)
afreq
perWetlands = 100 * sum(afreq[9:10,2])/sum(afreq[1:10,2])
perWetlands
```

### Convert polygons to raster ###
Things to make note of:

* CRS - Units and area are important!
* Grid size

```{r,message=F,warning=F,error=T,fig.keep='last'}
# Albers Equal Area Conic
AEA <- CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 
                                 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs") 
# Transform projection of NHD 
nhdAEA = spTransform(vilasNHD,AEA)


# Length and width of area in km 
width = (extent(nhdAEA)[4] - extent(nhdAEA)[3])/1000
length = (extent(nhdAEA)[2] - extent(nhdAEA)[1])/1000

# Create an empty raster 
r1 = raster(ncol=width*10, nrow=length*10) #approximately 100 m grid
r1
extent(r1) = extent(nhdAEA)
r1
```

We now have an empty raster that has the desired geometry 

```{r,message=F,warning=F,error=T,fig.keep='last'}
r2 = rasterize(nhdAEA,r1,field = 1, background = 0)
r2
plot(r2)
# Reclassify raster where all zero values are NA.
class1 <- reclassify(r2, cbind(0, NA))
class1
```

What would it look like with a 1km grid cell? 

```{r,message=F,warning=F,error=T,fig.keep='last'}
r1k = raster(ncol=width, nrow=length) #approximately 1 km grid
extent(r1k) = extent(nhdAEA)
r2k = rasterize(nhdAEA,r1k,field = 1, background = 0)
plot(r2k)
```

Can change the function of how cells are counted. 

`To determine what values to assign to cells that are covered by multiple spatial features.`
`You can use functions such as min, max, or mean, or one of the following character values:`
`'first', 'last', 'count'. The default value is 'last'.`

```{r,message=F,warning=F,error=T,fig.keep='last'}
r2kc = rasterize(nhdAEA,r1k,field = 1,background = 0, fun=function(x,...){length(x)})
plot(r2kc)
```
Note:

`You can also pass multiple functions using a statement like`
`fun=function(x, ...) c(length(x),mean(x)),` 
`in which case the returned object is a RasterBrick (multiple layers).`


## Homework ##

NHD data for Wisconsin:
http://prd-tnm.s3-website-us-west-2.amazonaws.com/?prefix=StagedProducts/Hydrography/NHD/State/HighResolution/Shape/


1) Calculate the percent water in each county in Wisconsin using the 2011 NLCD. 
2) Calculate the percent water in each county using the NHD. 




