---
title: "ZOO955 - Overlays"
author: "Hilary Dugan"
date: "2/22/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vector and Raster Data ##
It is common that you will have to use both vector and raster data in your work. Today we'll explore some common techniques for working with multiple data types. 

#### Lake point data ####
Read in lake point data from .csv 
```{r,message=F,warning=F,fig.width=3,fig.height=3}
library(sp)
library(raster)
library(rgdal)
lakes = read.csv('Data/nlakes.csv',stringsAsFactors = F)
coordinates(lakes) = ~Longitude+Latitude
crs(lakes) <- CRS("+proj=longlat +datum=WGS84")
plot(lakes,cex=2)
text(lakes,labels = lakes$Lake,cex=0.8)
```

#### Counties data ####
Read in lake point data from .csv 
```{r,message=F,warning=F}
library(rgdal)
counties = readOGR('C:/Users/hdugan/Dropbox/RandomR/GIS_Wisconsin/County_Boundaries_24K/County_Boundaries_24K.shp',layer='County_Boundaries_24K')
vilas = counties[counties@data$COUNTY_NAM == 'Vilas',]
plot(vilas)
```

#### National hydrography shapefiles ####
Read in lake point data from .csv 
```{r,message=F,warning=F}
library(rgdal)
vilasNHD = readOGR('Data/vilasNHD.shp',layer = 'vilasNHD')
plot(vilasNHD)
```

#### NLCD raster data ####
Read in lake point data from .csv 
```{r,message=F,warning=F}
library(rgdal)
nlcd = raster('Data/vilasNLCD.grd')
```
