---
title: "Zoo955 - Working With Shapefiles"
author: "Hilary Dugan"
date: "February 6, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Download data
Easy to search for 'spatial data' under NTL categories. Today we'll be working with Lake Mendota spatial data including the lake boundary, bathymetry/contours, and watershed shapefiles. 

*Watersheds*:
https://lter.limnology.wisc.edu/dataset/north-temperate-lakes-lter-yahara-lakes-district-lake-watersheds

*Bathymetry*: 
https://lter.limnology.wisc.edu/dataset/north-temperate-lakes-lter-yahara-lakes-district-bathymetry

*Lakes*:
https://lter.limnology.wisc.edu/dataset/north-temperate-lakes-lter-yahara-lakes-district-boundary

## Extract Lake Mendota data

Read in shapefiles using `readOGR`. Because there is an attached data.frame, you can use `stringsAsFactors = F` just like a normal data.frame
```{r lakes, message=FALSE,warning=FALSE}
library(rgdal)
lakes = readOGR('Data/yld_study_lakes.shp',layer = 'yld_study_lakes',stringsAsFactors = F)
summary(lakes)
```

We can see that there are many attributes associated with the shapefiles. Since we're interested in Lake Mendota, we can subset based on SHAIDNAME, LAKEID, or LAKE_NAME

```{r mendota}
mendota = lakes[lakes@data$LAKEID == 'ME',]
```

The watersheds and bathymetry are already subsetted, which is handy
```{r mendota.watershed}
mendota.ws = readOGR('Data/YaharaBasins/Mendota_Basin.shp',layer = 'Mendota_Basin',stringsAsFactors = F,verbose = F)
mendota.bathy = readOGR('Data/Bathymetry/mendota-contours-all.shp',layer = 'mendota-contours-all',verbose = F)
```

## SpatialPolygonsDataFrame Structure

The structure of `SpatialPolygonsDataFrame` make a lot more sense when you take a minute to look at the internal structure. It gets overwhelming when there are lots of polygons, so let's start with our Mendota object
```{r mendota.stucture}
str(mendota)
```

## Extract Properties

Knowing the structure makes extracting properties much easier. 
```{r area}
mendota@proj4string
mendota@polygons[[1]]@area
mendota@polygons[[1]]@Polygons[[1]]@area
```

## Use slot function 
Can also use the `slot` function. Usage: `slot(object, name)`

or `slotNames(x)` 

```{r slots}
slotNames(lakes)

slotNames(lakes@polygons[[1]]) # Can use within the lists as well 

slot(lakes,'proj4string')
```

Can nest within an `apply` function to extra all the properties 

```{r slot.apply}
sapply(slot(lakes, "polygons"), slot, "ID")
sapply(slot(lakes, "polygons"), slot, "area")
```

## Working with the bathymetry data

There are 88 polygons `nrow(mendota.bathy)`. Each depth interval has multiple polygons.

What do the polygons look like? 
```{r bathygons}
plot(mendota.bathy[10,],col='gold')
plot(mendota.bathy[25,],col='red',add=T)
```

Looking at `mendota.bathy@data`, there are only three columns **ID**, **DEPTH_FT**, and **DEPTH_M**

Would be handy to have the area of each polygon in the table. Can use our `sapply` function to extract the areas and add them to the data.frame

```{r bathygons.data}
mendota.bathy@data$AREA_m2 = sapply(slot(mendota.bathy, "polygons"), slot, "area")
head(mendota.bathy@data)
```

We can work with this data.frame the same we would with any data.frame
```{r bathygons.area, warning=FALSE, message=FALSE}
library(dplyr)
polygon.areas = mendota.bathy@data %>% group_by(ID) %>%
  summarise_at('AREA_m2',sum) %>%
  arrange(ID)
head(polygon.areas)
```

## But wait. 
Take a look at one individual depth. The polygons overlap. So we're overestimating areas. 
```{r bathygons.area2, warning=FALSE, message=FALSE}
thirty = mendota.bathy[mendota.bathy@data$ID == '30',]
sapply(slot(thirty, "polygons"), slot, "area") # Get areas

plot(thirty[1,],col='navy')
plot(thirty[2:6,],col='red',add=T)
```

We can create a **union** of these polygons by joining intersecting geometries
```{r bathygons.area3, warning=FALSE, message=FALSE}
library(rgeos)
union.thirty = gUnaryUnion(thirty)
plot(union.thirty)
sum(sapply(slot(union.thirty, "polygons"), slot, "area")) # new area
sum(sapply(slot(thirty, "polygons"), slot, "area")) # compare to old area
```
Now it would be nice if we could `group_by` to create a summarised data.frame with updated areas. But we can't, because dplyr won't read `sp` objects. Would have to do some crazy apply function, or a loop. 

## Welcome to sf
https://cran.r-project.org/web/packages/sf/vignettes/sf1.html

**Simple features** refer to a formal standard (ISO 19125-1:2004) that describes how objects in the real world can be represented in computers, with emphasis on the spatial geometry of these objects. It also describes how such objects can be stored in and retrieved from databases, and which geometrical operations should be defined for them.

The standard is widely implemented in spatial databases (such as PostGIS), commercial GIS (e.g., ESRI ArcGIS) and forms the vector data basis for libraries such as GDAL. A subset of simple features forms the GeoJSON standard.

R has well-supported classes for storing spatial data (sp) and interfacing to the above mentioned environments (rgdal, rgeos), but has so far lacked a complete implementation of simple features, making conversions at times convoluted, inefficient or incomplete. The package sf tries to fill this gap, and **aims at succeeding sp in the long term**.

```{r sf, warning=FALSE, message=FALSE}
library(sf)
sf.bathy <- st_read('Data/Bathymetry/mendota-contours-all.shp')
plot(sf.bathy)

lakes = st_read('Data/yld_study_lakes.shp')
mendota = lakes %>% filter(LAKEID == 'ME')
```

You *can* convert `sf` objections to `sp` objects: `as(mendota, "Spatial")`

## sf areas
Most geometries are WAY easier to calculate 
```{r sf.areas, warning=FALSE, message=FALSE}
head(st_area(sf.bathy))

areas = sf.bathy %>% mutate(AREA.m2 = st_area(sf.bathy)) %>%
  group_by(ID) %>%
  summarise(areas = st_union(geometry))

data.frame(ID = areas$ID,area = st_area(areas),old.areas = polygon.areas)
```

## sf CRS
sf ojects work with EPSG codes nicely, and prefer them over proj4strings. Mostly for convience. 

Coordinate reference systems can be applied using `st_crs`

Coordinate reference system transformations can be carried out using `st_transform`
```{r sf.crs, warning=FALSE, message=FALSE}
st_crs(areas) = st_crs(mendota)
#Could also do: st_crs(areas) <- 4326 (but need to know espg code)
```

# sf geometrical operations
The commands `st_intersects`, `st_disjoint`, `st_touches`,
`st_crosses`, `st_within`, `st_contains`, `st_overlaps`,
`st_equals`, `st_covers`, `st_covered_by`, `st_equals_exact` and
`st_is_within_distance` return a sparse matrix with matching (TRUE)
indexes, or a full logical matrix:
```{r}
st_intersects(areas,mendota)
```

The commands `st_buffer`, `st_boundary`, `st_convexhull`,
`st_union_cascaded`, `st_simplify`, `st_triangulate`,
`st_polygonize`, `st_centroid`, `st_segmentize`, and `st_union`
return new geometries. Example of st_buffer:

```{r}
buffer <- st_buffer(mendota, dist = 3000)
plot(buffer$geometry)
plot(mendota$geometry,add=T,col='turquoise')
```

Commands `st_intersection`, `st_union`, `st_difference`,
`st_sym_difference` return new geometries that are a function of
pairs of geometries:
```{r}
bathy.diff = st_difference(areas[1,]$geometry,areas[6,]$geometry)
plot(bathy.diff,col='turquoise')
```

## Homework

1) Define 7 of these using simple language (1 sentence if possible): `st_intersects`, `st_disjoint`, `st_touches`,`st_crosses`, `st_within`, `st_contains`, `st_overlaps`,
`st_equals`, `st_covers`, `st_covered_by`, `st_equals_exact` and
`st_is_within_distance`,`st_buffer`, `st_boundary`, `st_convexhull`,
`st_union_cascaded`, `st_simplify`, `st_triangulate`,
`st_polygonize`, `st_centroid`, `st_segmentize`, and `st_union`

Preferably choose ones that you don't already know. 

2) Make a 500 m buffer of the 4 southern LTER lakes. Which buffers overlap? 

3) [This question is considerably more difficult. Try your best, but don't spend more than 30 minutes on this.] Increase the size of the lakes by 2x. What is the percent of Mendota that overlaps with Monona? 


