---
title: "Zoo955 - Week 3"
author: "Hilary Dugan"
date: "2/13/2018"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1
Definefunctions from the `sf` package

`st_intersects` =

`st_disjoint` =

`st_touches` =

`st_crosses` =

`st_within` =

`st_contains` =

`st_overlaps` =

`st_equals` =

`st_covers` =

`st_covered_by` =

`st_equals_exact` =

`st_is_within_distance` =

`st_buffer` =

`st_boundary` =

`st_convexhull` =

`st_union_cascaded` =

`st_simplify` =

`st_triangulate` =

`st_polygonize` =

`st_centroid` =

`st_segmentize` =

`st_union` =

## Question 2 
**Make a 500 m buffer of the 4 southern LTER lakes. Which buffers overlap?**

*Lakes Data*:
https://lter.limnology.wisc.edu/dataset/north-temperate-lakes-lter-yahara-lakes-district-boundary

Read in shapefiles using `readOGR`
```{r lakes, message=FALSE,warning=FALSE}
library(sf)
lakes = st_read(dsn = '../Lecture3_Shapefiles/Data/yld_study_lakes.shp',stringsAsFactors = F)
```

## Question 3
Increase the size of the lakes by 2x. What is the percent of Mendota that overlaps with Monona?
