---
title: "Zoo955 - Week 5"
author: "Hilary Dugan"
date: "3/05/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use the 'Data/WI_CAVG_LatLong1.nc' file. In this lecture we dealt with the "temperature" variable. This file is the "climatology" variable. 

Find the metadata for these dataset. http://berkeleyearth.org/data/. We are using *Gridded Data*. *Monthly Land. Average Temperature (TAVG; 1753 – Recent)*. 

## 1) The climatology variable is a monthly average for each cell. What years does this average represent? 
January 1951 to December 1980 reported in degrees C

## 2) Plot the August averages for Wisconsin. 
```{r,message=F,warning=F}
library(raster)
br = brick('../Lecture5_RasterBricks/Data/WI_CAVG_LatLong1.nc') 
august = br[[8]] 
plot(august)
```

## 3) Extract the averages for the cell over Lake Mendota. Plot the monthly temperature averages. 
```{r,message=F,warning=F}
# Option 1
mendota = data.frame(lat = 43.1, long = -89.4)
coordinates(mendota) = ~long+lat
menTemp = extract(br,y = mendota)
plot(menTemp[1,],pch=16,ylab = 'Temperature (*C)',xlab = 'Month')
```

## 4) What is the August average for Lake Mendota? 
```{r,message=F,warning=F}
menTemp[1,8]
```
The august average temperature is 20.2
